---
title: 'Homework: Numerical linear algebra'
output:
  html_document:
    df_print: paged
  html_notebook:
    code_folding: none
    highlight: textmate
---

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: {
    Macros: {
      bA: "{\\boldsymbol{A}}",
      bx: "{\\boldsymbol{x}}",
      bb: "{\\boldsymbol{b}}"
    }
  }
});
</script>


```{r setup, include=FALSE}
source(file.path("..", "R", "util.R"))

required_packages <- c("microbenchmark")
install_and_load_packages(required_packages)
```


# Exercise 1: Comparing different numerical linear algebra algorithms for solving linear systems

In this exercise, we consider the problem of solving a linear system $\bA \bx = \bb$ for $\bx$.
We compare the three methods we learned in the class: LU, Cholesky, and QR decompositions.
(Of course, LU applies to more general systems and QR can be used to solve least squares, but here we focus on positive definite systems.)

## Part A: Racing for solution &mdash; speed comparison 

We first compare their computational speed. 
Fill in the code below and `bench::mark()` the three algorithms.

**Questions:**
What are relative speeds among the algorithms?
Do relative speeds agree with what you expect from the complexity analyses?
If not (quite), why might that be?

Solving via LU decomposition takes about 150 ms (median), solving via cholesky decomposition takes about 143 ms (median), solving via QR decomposition takes about 588 ms (median).

Theoretically, complexity of solving via LU decomposition is $\frac{2}{3}p^3+2p^2$, complexity of solving via Cholesky decomposition is $\frac{1}{3}p^3+2p^2$, complexity of solving via QR decomposition is $\frac{4}{3}p^3$. Roughly, solving via LU decomposition should be two times slower than solving via Cholesky decomposition. Solving via QR decomposition should be four times slower than solving via Cholesky decomposition. The relative speed of solving via cholesky decomposition and solving via QR decomposition is expected. The relative speed between solving via cholesky decomposition and solving via LU decomposition is not expected. The reason may be that R needs to allocate a new memory space for L matrix after Cholesky decomposition.

**Note:**
I misspoke about the R's built-in `chol()` function during the lecture:
when applied to a positive-definite `A`, the function actually returns an _upper-triangular_ matrix `R` such that `t(R) %*% R == A`.

```{r, eval=FALSE}
# Import the `rand_positive_def_matrix` function
source(file.path("R", "random_matrix.R"))

mat_size <- 1024L
A <- rand_positive_def_matrix(mat_size, cond_num = 1e3)
b <- rnorm(mat_size)

solve_via_cholesky <- function(A, b) {
  L = chol(A)
  x = backsolve(L, backsolve(L, b, transpose = TRUE))
  return(x)
}

bench::mark(
  solve(A, b)
)
bench::mark(
  solve_via_cholesky(A, b)
)
bench::mark(
  qr.solve(qr(A, LAPACK = TRUE), b)
)
```

## Part B: Competition in terms of numerical accuracy/stability

We now compare the three methods in terms of numerical accuracy/stability.
To this end, we set up the following simulation study. 
We first generate a "ground truth" solution vector $\bx_0$.
We then compute an "numerical-error-free" $\bb = \bA \bx_0$ by carrying out the matrix-vector multiplication using the `long double` type, which (on most hardware and compilers) provides [additional 12 bits of precision](https://en.wikipedia.org/wiki/Extended_precision#x86_extended_precision_format).
Of course, the vector $\bb$ computed as such still suffers from numerical errors, but the idea is that the numerical errors from this high-precision matrix-vector multiplication is much smaller than the errors caused by numerically solving $\bA \bx = \bb$ for $\bx$.
We can thus assess the accuracy of the three solvers by comparing the numerically computed $\bx$ to the ground truth $\bx_0$.

### Task &#x2F00;

First compare the outputs of matrix-vector multiplication $\bx \to \bA \bx$ using `double` and `long double` using the provided Rcpp functions.

**Questions:**
What is the relative difference in $\ell^2$-norm? 
How about the coordinate-wise relative differences?
Are the observed magnitudes of the differences what you'd expect?

The relative difference in $\ell^2$-norm is 5.6e-16 ($2^{-50.7}$). The coordinate-wise relative differences is 9.5e-13 ($2^{-40}$). The observed magnitudes of the differences is what I expect. 'long double' provides 12 more digits precision than 'double' type. The first 52 digits should be the same. The relative difference two forms should be around $2^{-52}$ 
```{r, eval=FALSE}
Rcpp::sourceCpp(file.path("src", "matvec_double.cpp"))
Rcpp::sourceCpp(file.path("src", "matvec_ldouble.cpp"))

set.seed(1918)

mat_size <- 1024L
A <- rand_positive_def_matrix(mat_size, cond_num = 1e3)
x <- rnorm(mat_size)

Ax_double = matvec_double(A, x)
Ax_ldouble = matvec_ldouble(A, x)

rel_diff_l2 = function(x0, x1){
  return(sqrt(sum((x0 - x1) ^ 2)) / sqrt(sum(x0 ^ 2)))
}
rel_diff_coordinate = function(x0, x1){
  return(abs(x0 - x1) / abs(x0))
}
Ax_rel_diff_l2 = rel_diff_l2(Ax_ldouble, Ax_double)
Ax_rel_diff_coordinate = rel_diff_coordinate(Ax_ldouble, Ax_double)
max(Ax_rel_diff_coordinate)
```

### Task &#x2F06;

Now randomly generate $\bA$ so that its condition number is $10^6$.
Then solve a positive-definite system $\bA \bx = \bb$ for $\bx$ using the three algorithms and compare their outputs to the ground truth $\bx_0$.

**Questions:**
Which algorithm appears to be more accurate than the others? 
Visually demonstrate your answer.

The LU decomposition appears to be more accurate than others.
```{r, eval=FALSE}
set.seed(1918)
cond_num <- 1e6

# Larger matrices could incur substantial computational time under base R BLAS
mat_size <- 1024L 

A <- rand_positive_def_matrix(mat_size, cond_num)
x <- rnorm(mat_size)

b = matvec_ldouble(A, x)
x_lu = solve(A, b)
x_chol = solve_via_cholesky(A, b)
x_qr = qr.solve(qr(A, LAPACK = TRUE), b)

lu_diff = rel_diff_coordinate(x_lu, x)
chol_diff = rel_diff_coordinate(x_chol, x)
qr_diff = rel_diff_coordinate(x_qr, x)

result_df = data.frame(method = c(rep('LU', mat_size), rep('Cholesky', mat_size), rep('QR', mat_size)), diff = log(c(lu_diff, chol_diff, qr_diff)))
boxplot(diff ~ method, data = result_df, ylab = "relative coordinate difference in log10 scale")
```

### Task &#x4E09;

In Task &#x2F06;, we compared the three algorithms in one randomly generated example.
Now we consider a more systematic (though hardly comprehensive) comparison via repeated simulations.
We also vary the condition number of $\bA$ and assess whether the results would hold across varying degrees of ill-conditioning.

**Questions/To-do's:**

* Using the starter code provided, calculate various summary measures of the numerical errors.
* Integrate into the provided code one another (or more, if you like) meaningful metric(s) of your choice to summarize the numerical error.
* Visually explore how the three algorithms compare with each other in their accuracy. See if different error metrics tell different stories; they might or might not.
* Vary the condition number in the range $10^6 \sim 10^{12}$, e.g. by trying $10^6$, $10^9$, and $10^{12}$.
* Do you see any patterns in the numerical errors across the three algorithms, metrics, and/or condition numbers? Show some plots to support your conclusion.

When condition number is $10^6$ and $10^9$, QR has more numerical error in all metrics compared with other two methods and Cholesky has the smallest error in all three methods under all metrics. When condition number is $10^12$, QR method has the similar results as LU and Cholesky has still the smallest error.

As the condition number increase, the numerical error also increase and linearly in log scale. We can see this from the scale of the y-axis.
**Note:** 
The QR solver will throw an error when the system is ill-condition enough that the numerical solution might not be very accurate. 
To force it to return the solution in any case, set `tol = .Machine$double.eps`.

```{r, eval=FALSE}
# Utility functions for bookkeeping simulation results.
source(file.path("R", "num_linalg_sim_study_helper.R"))

n_sim <- 32L
mat_size <- 512L
cond_num <- 1e6
metrics <- c("norm", "median", "five_percentile", "ninety_five_percentile", "mean")

rel_error_list <- lapply(
  c("lu", "chol", "qr"), 
  function(method) pre_allocate_error_list(n_sim, metrics)
)

for (sim_index in 1:n_sim) {
  A <- rand_positive_def_matrix(mat_size, cond_num)
  x <- rnorm(mat_size) 
  b <- matvec_ldouble(A, x)
  x_approx <- list( 
    lu = solve(A, b),
    chol = solve_via_cholesky(A, b),
    qr = qr.solve(qr(A, LAPACK = TRUE), b, tol=.Machine$double.eps)
  )
  for (method in c("lu", "chol", "qr")) {
    rel_error <- lapply(
      metrics, 
      function (metric) calc_rel_error(x, x_approx[[method]], metric)
    )
    names(rel_error) <- metrics
    for (metric in names(rel_error)) {
      rel_error_list[[method]][[metric]][sim_index] <- rel_error[[metric]]
    }
  }
}

result_df = data.frame(method = c(rep('LU', n_sim), rep('Cholesky', n_sim), rep('QR', n_sim)), diff = log(c(rel_error_list$lu$norm, rel_error_list$chol$norm, rel_error_list$qr$norm)))
boxplot(diff ~ method, data = result_df, ylab = "l2 norm difference in log10 scale")

result_df = data.frame(method = c(rep('LU', n_sim), rep('Cholesky', n_sim), rep('QR', n_sim)), diff = log(c(rel_error_list$lu$median, rel_error_list$chol$median, rel_error_list$qr$median)))
boxplot(diff ~ method, data = result_df, ylab = "median of coordinate difference in log10 scale")

result_df = data.frame(method = c(rep('LU', n_sim), rep('Cholesky', n_sim), rep('QR', n_sim)), diff = log(c(rel_error_list$lu$five_percentile, rel_error_list$chol$five_percentile, rel_error_list$qr$five_percentile)))
boxplot(diff ~ method, data = result_df, ylab = "five_percentile of coordinate difference in log10 scale")

result_df = data.frame(method = c(rep('LU', n_sim), rep('Cholesky', n_sim), rep('QR', n_sim)), diff = log(c(rel_error_list$lu$ninety_five_percentile, rel_error_list$chol$ninety_five_percentile, rel_error_list$qr$ninety_five_percentile)))
boxplot(diff ~ method, data = result_df, ylab = "ninety_five_percentile of coordinate difference in log10 scale")

result_df = data.frame(method = c(rep('LU', n_sim), rep('Cholesky', n_sim), rep('QR', n_sim)), diff = log(c(rel_error_list$lu$mean, rel_error_list$chol$mean, rel_error_list$qr$mean)))
boxplot(diff ~ method, data = result_df, ylab = "mean of coordinate difference in log10 scale")


cond_num <- 1e9
rel_error_list <- lapply(
  c("lu", "chol", "qr"), 
  function(method) pre_allocate_error_list(n_sim, metrics)
)

for (sim_index in 1:n_sim) {
  A <- rand_positive_def_matrix(mat_size, cond_num)
  x <- rnorm(mat_size) 
  b <- matvec_ldouble(A, x)
  x_approx <- list( 
    lu = solve(A, b),
    chol = solve_via_cholesky(A, b),
    qr = qr.solve(qr(A, LAPACK = TRUE), b, tol=.Machine$double.eps)
  )
  for (method in c("lu", "chol", "qr")) {
    rel_error <- lapply(
      metrics, 
      function (metric) calc_rel_error(x, x_approx[[method]], metric)
    )
    names(rel_error) <- metrics
    for (metric in names(rel_error)) {
      rel_error_list[[method]][[metric]][sim_index] <- rel_error[[metric]]
    }
  }
}

result_df = data.frame(method = c(rep('LU', n_sim), rep('Cholesky', n_sim), rep('QR', n_sim)), diff = log(c(rel_error_list$lu$norm, rel_error_list$chol$norm, rel_error_list$qr$norm)))
boxplot(diff ~ method, data = result_df, ylab = "l2 norm difference in log10 scale")

result_df = data.frame(method = c(rep('LU', n_sim), rep('Cholesky', n_sim), rep('QR', n_sim)), diff = log(c(rel_error_list$lu$median, rel_error_list$chol$median, rel_error_list$qr$median)))
boxplot(diff ~ method, data = result_df, ylab = "median of coordinate difference in log10 scale")

result_df = data.frame(method = c(rep('LU', n_sim), rep('Cholesky', n_sim), rep('QR', n_sim)), diff = log(c(rel_error_list$lu$five_percentile, rel_error_list$chol$five_percentile, rel_error_list$qr$five_percentile)))
boxplot(diff ~ method, data = result_df, ylab = "five_percentile of coordinate difference in log10 scale")

result_df = data.frame(method = c(rep('LU', n_sim), rep('Cholesky', n_sim), rep('QR', n_sim)), diff = log(c(rel_error_list$lu$ninety_five_percentile, rel_error_list$chol$ninety_five_percentile, rel_error_list$qr$ninety_five_percentile)))
boxplot(diff ~ method, data = result_df, ylab = "ninety_five_percentile of coordinate difference in log10 scale")

result_df = data.frame(method = c(rep('LU', n_sim), rep('Cholesky', n_sim), rep('QR', n_sim)), diff = log(c(rel_error_list$lu$mean, rel_error_list$chol$mean, rel_error_list$qr$mean)))
boxplot(diff ~ method, data = result_df, ylab = "mean of coordinate difference in log10 scale")




cond_num <- 1e12
rel_error_list <- lapply(
  c("lu", "chol", "qr"), 
  function(method) pre_allocate_error_list(n_sim, metrics)
)

for (sim_index in 1:n_sim) {
  A <- rand_positive_def_matrix(mat_size, cond_num)
  x <- rnorm(mat_size) 
  b <- matvec_ldouble(A, x)
  x_approx <- list( 
    lu = solve(A, b),
    chol = solve_via_cholesky(A, b),
    qr = qr.solve(qr(A, LAPACK = TRUE), b, tol=.Machine$double.eps)
  )
  for (method in c("lu", "chol", "qr")) {
    rel_error <- lapply(
      metrics, 
      function (metric) calc_rel_error(x, x_approx[[method]], metric)
    )
    names(rel_error) <- metrics
    for (metric in names(rel_error)) {
      rel_error_list[[method]][[metric]][sim_index] <- rel_error[[metric]]
    }
  }
}

result_df = data.frame(method = c(rep('LU', n_sim), rep('Cholesky', n_sim), rep('QR', n_sim)), diff = log(c(rel_error_list$lu$norm, rel_error_list$chol$norm, rel_error_list$qr$norm)))
boxplot(diff ~ method, data = result_df, ylab = "l2 norm difference in log10 scale")

result_df = data.frame(method = c(rep('LU', n_sim), rep('Cholesky', n_sim), rep('QR', n_sim)), diff = log(c(rel_error_list$lu$median, rel_error_list$chol$median, rel_error_list$qr$median)))
boxplot(diff ~ method, data = result_df, ylab = "median of coordinate difference in log10 scale")

result_df = data.frame(method = c(rep('LU', n_sim), rep('Cholesky', n_sim), rep('QR', n_sim)), diff = log(c(rel_error_list$lu$five_percentile, rel_error_list$chol$five_percentile, rel_error_list$qr$five_percentile)))
boxplot(diff ~ method, data = result_df, ylab = "five_percentile of coordinate difference in log10 scale")

result_df = data.frame(method = c(rep('LU', n_sim), rep('Cholesky', n_sim), rep('QR', n_sim)), diff = log(c(rel_error_list$lu$ninety_five_percentile, rel_error_list$chol$ninety_five_percentile, rel_error_list$qr$ninety_five_percentile)))
boxplot(diff ~ method, data = result_df, ylab = "ninety_five_percentile of coordinate difference in log10 scale")

result_df = data.frame(method = c(rep('LU', n_sim), rep('Cholesky', n_sim), rep('QR', n_sim)), diff = log(c(rel_error_list$lu$mean, rel_error_list$chol$mean, rel_error_list$qr$mean)))
boxplot(diff ~ method, data = result_df, ylab = "mean of coordinate difference in log10 scale")

```
